Automatic Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Handlebar Perturbations\\Version 2.dev

### 2024-09-28 main.tex
Addressed some of Riender's feedback.

Turning the balance assist system on significantly \((p<0.05)\) reduces the
odds that a perturbation results in a fall while cycling at a speed of
6~\si{\kph}. Figure~\ref{fig:probability-6kph} visualizes the probability of
falling as a function of the mean and centered angular impulse per participant
for the balance assist state on and off while keeping all other explanatory
factors at their centered mean values. This figure is created by setting all
explanatory variables to their mean and calculating the probability from
Equation~\ref{eq:log-regress} for only change in angular impulse given the
estimates in Table~\ref{tab:freq-coefs-6}. The table indicates that the balance
assist system halves (0.53) the odds that a perturbation results in a fall.
This figure shows that for relatively large impulses the probability to fall is
unity for both states of balance assist on and off. And for relatively small
impulses the probability to fall is null for both states. But for impulses in
the magnitude region (about -1 to 0.5 STD), i.e. around the mean-centered
perturbation resistance threshold, the probability of falling is significantly
lowered with the balance assist system on. The skewness of the probability
curves arrives from the interaction effects. Figure~\ref{fig:probability-10kph}
shows the same result for the 10~\si{\kph} trials which has a similar trend of
reducing the probability to fall with the balance assist system turned on, but
the effect is not significant.
%
\begin{figure}
  \centering
  \subcaptionbox{Fall probability for the 6~\si{\kph} trials.
    \label{fig:probability-6kph}}{
    \includegraphics[width=80mm]{figures/predicted_fall_probability_6kmh.png}
  }
  \subcaptionbox{Fall probability for the 10~\si{\kph} trials.
    \label{fig:probability-10kph}}{
    \includegraphics[width=80mm]{figures/predicted_fall_probability_10kmh.png}
  }
  \caption{Comparison of predicted fall probability for balance assist on
    (solid) or off (dashed) when all other predictor variables are set to zero,
    except for the interaction effect between the balance assist and the
    angular impulse. The abscissa is the standard deviation around the mean
    (all perturbations), centered per participant.}
  \label{fig:probability}
\end{figure}

@@

@@

\section{Conclusion}
%
Automatically controlling a steering motor in a bicycle using roll rate
feedback lowers the speed at which the bicycle is stable. This makes the
bicycle's low speed dynamics more akin to its uncontrolled high speed dynamics,
which is easier for a rider to control in balance. Perturbation forces applied
to the handlebar can cause a rider to fall and every rider has a perturbation
magnitude threshold at which they are more likely to fall than not. The
probability of falling when mechanically perturbed is significantly reduced
when traveling at 6~\si{\kph} when the balance assisting control is activated.
This effect is present when traveling at 10~\si{\kph} but more investigation is
needed to determine if the effect can be significant. The positive effect to
balance is rider independent and most effective in the regime of perturbations
near the limits of the rider's control authority. Given that similar effects
cause falls during bicycling, use of the balance assist system in real world
use cases will reduce the number of falls at low bicycling speeds.

\section*{Acknowledgements}
%
This study follows and draws from experimental and analysis methods originally
developed in unpublished research by Marco Reijne. The authors acknowledge
Felix Dauer, David Gabriel, Sierd Heida, Oliver Maier, Maarten Pelgrim, Marco
Reijne, and Arend Schwab for contributions to the development of the balance
assist bicycle. We acknowledge Shannon van de Velde and Marco Reijne's
development of the bicycle perturbation system. Allen Downey provided useful
insight in the statistical condsiderations. We thank all of the assistants that
helped run the experiments and the participants for their time an enthusiasm.
We thank Peter Stahlecker and Benjamin Gonzalez for valuable feedback in early
revisions.

>>>>

Turning the balance assist system on significantly \((p<0.05)\) reduces the
odds that a perturbation results in a fall while cycling at a speed of
6~\si{\kph}. Figure~\ref{fig:probability-6kph} visualizes the probability of
falling as a function of the mean and centered angular impulse per participant
for the balance assist state on and off while keeping all other explanatory
factors at their centered mean values. This figure is created by setting all
explanatory variables to their mean and calculating the probability from
Equation~\ref{eq:log-regress} for only change in angular impulse given the
estimates in Table~\ref{tab:freq-coefs-6}. The table indicates that the balance
assist system halves (0.53) the odds that a perturbation results in a fall.
This figure shows that for relatively large impulses the probability to fall is
unity for both states of balance assist on and off. And for relatively small
impulses the probability to fall is null for both states. But for impulses in
the magnitude region (about -1 to 0.5 STD), i.e. around the mean-centered
perturbation resistance threshold, the probability of falling is significantly
lowered with the balance assist system on. The skewness of the probability
curves arrives from the interaction effects. Figure~\ref{fig:probability-10kph}
shows the same result for the 10~\si{\kph} trials which has a similar trend of
reducing the probability to fall with the balance assist system turned on, but
the effect is not significant.
%
\begin{figure}
  \centering
  \subcaptionbox{Fall probability for the 6~\si{\kph} trials.
    \label{fig:probability-6kph}}{
    \includegraphics[width=80mm]{figures/predicted_fall_probability_6kmh.png}
  }
  \subcaptionbox{Fall probability for the 10~\si{\kph} trials.
    \label{fig:probability-10kph}}{
    \includegraphics[width=80mm]{figures/predicted_fall_probability_10kmh.png}
  }
  \caption{Comparison of predicted fall probability for balance assist on
    (solid) or off (dashed) when all other predictor variables are set to zero,
    except for the interaction effect between the balance assist and the
    angular impulse. The abscissa is the standard deviation around the mean
    of all perturbations, centered per participant.}
  \label{fig:probability}
\end{figure}

@@

\subsection{On Rider Bicycling Skill and Experience}
%
Each individual participant had their own perturbation resistance threshold,
i.e. the perturbation magnitude defining whether they would fall 50\% of the
time.
The value of this threshold is likely correlated with the skill and experience
of the participant at balancing a bicycle.
With this study, we cannot answer whether the balance assist system helps low
skilled participants more than highly skilled participants because that is
simply a function of how often a perturbation in a natural setting exceeds an
individual's threshold.
But, we did find that the system can help regardless of a participant's
individual balance performance.
On average, the balance assist system pushes the perturbation resistance
threshold up for all participants.
The vast majority of perturbations experienced in natural bicycling are below
an individual's perturbation resistance threshold, otherwise we would observe
bicyclists falling much more but low skilled riders may experience more
perturbations that exceed their threshold.
Determining how often certain perturbation types and magnitudes are experienced
in natural bicycling is an open research need and would help understand whether
a balance assist intervention would have a strong effect when used in the wild.

@@

\section{Conclusion}
%
Automatically controlling a steering motor in a bicycle using roll rate
feedback lowers the speed at which the bicycle is stable. This makes the
bicycle's low speed dynamics more akin to its uncontrolled high speed dynamics,
which is easier for a rider to control in balance. Perturbation forces applied
to the handlebar can cause a rider to fall and every rider has a perturbation
magnitude threshold at which they are more likely to fall than not. The
probability of falling when mechanically perturbed is significantly
(\(p=0.02\)) reduced when traveling at 6~\si{\kph} when the balance assisting
control is activated. This effect is present when traveling at 10~\si{\kph}
and is very close to significance (\(p=0.07\)) but more investigation is needed
to determine if the effect can be significant. The positive effect to balance
is rider independent and most effective in the regime of perturbations near the
limits of the rider's control authority. Given that similar effects cause falls
during bicycling, use of the balance assist system in real world use cases will
reduce the number of falls at low bicycling speeds.

\section*{Acknowledgements}
%
This study follows and draws from experimental and analysis methods originally
developed in unpublished research by Marco Reijne. The authors acknowledge
Felix Dauer, David Gabriel, Sierd Heida, Oliver Maier, Maarten Pelgrim, Marco
Reijne, and Arend Schwab for contributions to the development of the balance
assist bicycle. We acknowledge Shannon van de Velde and Marco Reijne's
development of the bicycle perturbation system. Allen Downey provided useful
insight in the statistical condsiderations. We thank all of the assistants that
helped run the experiments and the participants for their time an enthusiasm.
We thank Peter Stahlecker, Benjamin Gonzalez, and Riender Happee for valuable
feedback in early revisions.
